Bird beak's elaboration and innovation

### 2022-05-26 Analysis/04-Figure1_with_ellipse_scalling_and_8D_histograms.Rmd
Update 04-Figure1_with_ellipse_scalling_and_8D_histograms.Rmd

```{r}
box_size <- 1.5
pdf(file = "../Manuscript/Figures/Figure1.pdf", height = box_size*8, width = box_size*10)

@@

## PLOTTING THE SUPERORDERS
## Getting the tip order for the ladderised super order tree
tip_order <- c(
    "Paleognathae" = 1,
    "Galloanserae" = 2,
    "Otidimorphae" = 5,
    "Aequornithes" = 6,
    "Mirandornithes" = 3,
    "Columbimorphae" = 4,
    "Telluraves" = 8,
    "Strisores" = 7)

@@

## Plotting the ellipses
for(i in 1:length(tip_order)) {
    ## Select the colour and name
    name.col <- tip_colours[i]

    ## Plot the ellipse
    par(bty = "n", mar = c(1, 1, 1, 1)+0.1)
    plot.one.ellipse(superorder_shapespace, name.col, x = FALSE, y = FALSE, with.phylo = TRUE)

    ## Calculate which dimension contains the most variance
    add.dims(superorder_shapespace, name.col)
    n <- 1000
    covars_cent_tend <- lapply(sample.n(get.subsets(superorder_shapespace, subset = names(name.col))$covar, n), VCV.cent.tend, mean)
    dim_var <- apply((get.one.axis(covars_cent_tend[[1]], axis = 1, level = 0.95)), 2, dist)
    text_main <- paste0(names(name.col))
    text(0,0.3, text_main, cex = 1.2, col = name.col)

    ## Plot the elaboration/exploration
    par(mar = c(2, 1, 1, 1)+0.1)
    plot.one.proj.rej(results$super_results_between, name.col, x = TRUE, y = FALSE)
}

## PLOTTING THE ORDERS
## Matching the order and the colours to the ladderised tree
tip_order <- c(
    "Tinamiformes"= 1,
    "Galliformes"= 3,
    "Anseriformes"= 2,
    "Otidiformes"= 7,
    "Cuculiformes" = 8,
    "Gruiformes" = 9, 
    "Musophagiformes" = 10,
    "Procellariiformes" = 12,
    "Sphenisciformes" = 11,
    "Ciconiiformes" = 13,
    "Pelecaniformes"= 15,
    "Suliformes" = 14,
    "Podicipediformes" = 4,
    "Columbiformes" = 6,
    "Pterocliformes" = 5,
    "Charadriiformes" = 18,
    "Trogoniformes" = 21,
    "Bucerotiformes" = 22,
    "Coraciiformes" = 23,
    "Piciformes" = 24,
    "Strigiformes" = 20,
    "Accipitriformes" = 19, 
    "Falconiformes" = 25,
    "Psittaciformes" = 26,
    "Passeriformes" = 27,
    "Caprimulgiformes" = 16, 
    "Apodiformes" = 17)

@@

for(i in 1:length(tip_order)) {
    ## Select the colour and name
    name.col <- tip_colours[i]

    ## Plot the ellipse
    par(bty = "n", mar = c(1, 1, 1, 1)+0.1)
    plot.one.ellipse(order_shapespace, name.col, x = FALSE, y = FALSE, with.phylo = TRUE)

    ## Calculate which dimension contains the most variance
    add.dims(order_shapespace, name.col)
    n <- 1000
    covars_cent_tend <- lapply(sample.n(get.subsets(order_shapespace, subset = names(name.col))$covar, n), VCV.cent.tend, mean)
    dim_var <- apply((get.one.axis(covars_cent_tend[[1]], axis = 1, level = 0.95)), 2, dist)
    text_main <- paste0(names(name.col))
    text(0,0.3, text_main, cex = 1.2, col = name.col)

    ## Plot the elaboration/exploration
    par(bty = "n", mar = c(2, 1, 1, 1)+0.1)
    plot.one.proj.rej(results$order_results_between, name.col, x = TRUE, y = FALSE)
}
dev.off()

@@

# Passeriformes bit

@@

family_shapespace <- MCMCglmm.subsets(n = 1000, # Just run it on 1k matrices for now
      data          = shapespace,
      posteriors    = shapespace_passeriformes,
      group         = MCMCglmm.levels(shapespace_passeriformes)[c(famil_lvl, phylo_lvl)],
      rename.groups = c(levels(shapespace$level2)[-1], "phylogeny")) #[-1] is because the first level is ""
## Clean garbage
rm(shapespace_passeriformes)
```

First we can visualise the global bird shape space (or at least the 2 first of the 8 dimensions):

```{r, echo = FALSE}
## Quick load
load(file = "../Data/Processed/shapespace_passeriformes_suborder_family_results_list.rda")
results <- shapespace_passeriformes_suborder_family_results_list
rm(shapespace_passeriformes_suborder_family_results_list)
```

@@

```{r}

source("../Functions/covar.utilities.R")
source("../Functions/ploting.utilities.R")

box_size <- 1.5
pdf(file = "../Manuscript/Figures/Figure1_passeriformes.pdf", height = box_size*7, width = box_size*10)

## Global window
lay_mat_family <- rbind(
        matrix(c(1,1, 2:9), ncol = 10, nrow = 1), # line 1
        matrix(c(1,1, 10:17), ncol = 10, nrow = 1), # line 2
        matrix(c(1,1, 18:25), ncol = 10, nrow = 1), # line 3
        matrix(c(26:35), ncol = 10, nrow = 1), # line 4
        matrix(c(36:45), ncol = 10, nrow = 1), # line 5
        matrix(c(46:55), ncol = 10, nrow = 1) # line 6
    )
lay_mat_sub <- rbind(
        matrix(c(1,1, 2:5, rep(0, 4)), ncol = 10, nrow = 1) # line 7
    )

heights_vals <- rep(1, 8)
widths_vals <- rep(c(1, 1/3), 5)
layout_matrix <- layout(rbind(lay_mat_sub, (lay_mat_family+5)), widths = widths_vals, heights = heights_vals, respect = TRUE)

# layout.show(layout_matrix)

## TODO: FUNCTIONALISE THE TREE PLOTTING: START

## PLOTTING THE SUBORDERS
tree <- ladderize(tree)
## Get the tip_values
families_lvl <- shapespace$level2
suborders_lvl <- shapespace$level1

## Select the level
selected_level <- coloured_levels <- suborders_lvl

## Replace the order levels as colours
colour_palette <- c("grey", colour.palette(length(levels(coloured_levels)[-1])))
levels(coloured_levels) <- c("grey", colour.palette(length(levels(coloured_levels)[-1])))

## Shorten the tree
selected_tips <- character()
for(one_level in 2:length(levels(selected_level))) {
    selected_tips[one_level-1] <- rownames(shapespace)[which(selected_level == levels(selected_level)[one_level])[1]]
    names(selected_tips)[one_level-1] <- levels(selected_level)[one_level]
}

## Shorten the tree
tree_short <- drop.tip(tree, tip = tree$tip.label[!(tree$tip.label %in% selected_tips)])
tree_short$tip.label <- names(selected_tips)[match(tree_short$tip.label, selected_tips)]

## Setting the tips edges colours
tip_edges <- match.tip.edge(tree_short$tip.label, tree_short, replace.na = "grey")
## Rainbow the tip edges
tip_edges[!(tip_edges %in% "grey")] <- rev(colour.palette(Ntip(tree_short)))
## Adding the number of tips
tree_plot <- tree_short
n_species <- table(selected_level)
for(i in 1:Ntip(tree_plot)) {
    ## Get the number of species
    n_sp <- n_species[which(names(n_species) %in% tree_plot$tip.label[i])]
    tree_plot$tip.label[i] <- paste0(tree_plot$tip.label[i], " (", n_sp, " sp)")
}

## Plotting the tree
par(mar = c(0, 0, 0, 0)+0.1)
plot(tree_plot, show.tip.label = TRUE, edge.color = tip_edges, edge.width = 3, cex = 1)

## TODO: FUNCTIONALISE THE TREE PLOTTING: END

## Matching the order and the colours to the ladderised tree
tip_order <- c(
    "Tyranni" = 1,
    "Passeri" = 2)

tip_colours <- colour.palette(Ntip(tree_short))
names(tip_colours) <- names(sort(tip_order))
tip_colours_suborders <- tip_colours
save(tip_colours_suborders, file = "../Data/Processed/tip_colours_suborders.rda")

## Plotting the ellipses for each suborder
for(i in 1:length(tip_order)) {

    ## TODO: FUNCTIONALISE THE ELLIPSE PLOTTING: START

    ## Select the colour and name
    name.col <- tip_colours[i]

    ## Plot the ellipse
    par(bty = "n", mar = c(1, 1, 1, 1)+0.1)
    plot.one.ellipse(suborder_shapespace, name.col, x = FALSE, y = FALSE, with.phylo = TRUE)

    ## Calculate which dimension contains the most variance
    add.dims(suborder_shapespace, name.col)
    n <- 1000
    covars_cent_tend <- lapply(sample.n(get.subsets(suborder_shapespace, subset = names(name.col))$covar, n), VCV.cent.tend, mean)
    dim_var <- apply((get.one.axis(covars_cent_tend[[1]], axis = 1, level = 0.95)), 2, dist)
    text_main <- paste0(names(name.col))
    text(0,0.3, text_main, cex = 1.2, col = name.col)

    ## Plot the elaboration/exploration
    par(mar = c(2, 1, 1, 1)+0.1)
    plot.one.proj.rej(results$suborder_results_between, name.col, x = TRUE, y = FALSE)

        ## TODO: FUNCTIONALISE THE ELLIPSE PLOTTING: END
}

## PLOTTING THE FAMILIES
tree <- ladderize(tree)
## Select the level
selected_level <- coloured_levels <- families_lvl

## Replace the order levels as colours
colour_palette <- c("grey", colour.palette(length(levels(coloured_levels)[-1])))
levels(coloured_levels) <- c("grey", colour.palette(length(levels(coloured_levels)[-1])))

## Shorten the tree
selected_tips <- character()
for(one_level in 2:length(levels(selected_level))) {
    selected_tips[one_level-1] <- rownames(shapespace)[which(selected_level == levels(selected_level)[one_level])[1]]
    names(selected_tips)[one_level-1] <- levels(selected_level)[one_level]
}

## Shorten the tree
tree_short <- drop.tip(tree, tip = tree$tip.label[!(tree$tip.label %in% selected_tips)])
tree_short$tip.label <- names(selected_tips)[match(tree_short$tip.label, selected_tips)]

## Setting the tips edges colours
tip_edges <- match.tip.edge(tree_short$tip.label, tree_short, replace.na = "grey")
## Rainbow the tip edges
tip_edges[!(tip_edges %in% "grey")] <- rev(colour.palette(Ntip(tree_short)))

## Adding the number of tips
tree_plot <- tree_short
n_species <- table(selected_level)
for(i in 1:Ntip(tree_plot)) {
    ## Get the number of species
    n_sp <- n_species[which(names(n_species) %in% tree_plot$tip.label[i])]
    tree_plot$tip.label[i] <- paste0(tree_plot$tip.label[i], " (", n_sp, " sp)")
}

## Plotting the tree
par(mar = c(0, 0, 0, 0)+0.1)
plot(tree_plot, show.tip.label = TRUE, edge.color = tip_edges, edge.width = 3, cex = 1)

@@

tip_colours <- colour.palette(Ntip(tree_short))
names(tip_colours) <- names(sort(tip_order))
tip_colours_families <- tip_colours
save(tip_colours_families, file = "../Data/Processed/tip_colours_families.rda")

for(i in 1:length(tip_order)) {
    ## Select the colour and name
    name.col <- tip_colours[i]

    ## Plot the ellipse
    par(bty = "n", mar = c(1, 1, 1, 1)+0.1)
    plot.one.ellipse(family_shapespace, name.col, x = FALSE, y = FALSE, with.phylo = TRUE)

    ## Calculate which dimension contains the most variance
    add.dims(family_shapespace, name.col)
    n <- 1000
    covars_cent_tend <- lapply(sample.n(get.subsets(family_shapespace, subset = names(name.col))$covar, n), VCV.cent.tend, mean)
    dim_var <- apply((get.one.axis(covars_cent_tend[[1]], axis = 1, level = 0.95)), 2, dist)
    text_main <- paste0(names(name.col))
    text(0,0.3, text_main, cex = 1.2, col = name.col)

    ## Plot the elaboration/exploration
    par(bty = "n", mar = c(2, 1, 1, 1)+0.1)
    plot.one.proj.rej(results$family_results_between, name.col, x = TRUE, y = FALSE)
}
dev.off()

>>>>

```{r, eval = TRUE, echo = FALSE}
box_size <- 1.5
pdf(file = "../Manuscript/Figures/Figure1.pdf", height = box_size*8, width = box_size*10)

## SETTING THE WINDOW PARAMETERS

@@

## PLOTTING THE SUPERORDERS

## Getting the tip order for the ladderised super order tree
tip_order <- c(
    "Paleognathae" = 1,
    "Galloanserae" = 2,
    "Otidimorphae" = 5,
    "Aequornithes" = 6,
    "Mirandornithes" = 3,
    "Columbimorphae" = 4,
    "Telluraves" = 8,
    "Strisores" = 7)

@@

## Plotting the ellipses
for(i in 1:length(tip_order)) {
    ## Plot each pannel
    wrap.plot.ellipses(i, tip_colours_super_orders, superorder_shapespace, results$super_results_between)
}

## PLOTTING THE ORDERS

## Matching the order and the colours to the ladderised tree
tip_order <- c(
    "Tinamiformes"= 1,
    "Galliformes"= 3,
    "Anseriformes"= 2,
    "Otidiformes"= 7,
    "Cuculiformes" = 8,
    "Gruiformes" = 9, 
    "Musophagiformes" = 10,
    "Procellariiformes" = 12,
    "Sphenisciformes" = 11,
    "Ciconiiformes" = 13,
    "Pelecaniformes"= 15,
    "Suliformes" = 14,
    "Podicipediformes" = 4,
    "Columbiformes" = 6,
    "Pterocliformes" = 5,
    "Charadriiformes" = 18,
    "Trogoniformes" = 21,
    "Bucerotiformes" = 22,
    "Coraciiformes" = 23,
    "Piciformes" = 24,
    "Strigiformes" = 20,
    "Accipitriformes" = 19, 
    "Falconiformes" = 25,
    "Psittaciformes" = 26,
    "Passeriformes" = 27,
    "Caprimulgiformes" = 16, 
    "Apodiformes" = 17)

@@

## Plotting the ellipses
for(i in 1:length(tip_order)) {
    ## Plot each pannel
    wrap.plot.ellipses(i, tip_colours_orders, order_shapespace, results$order_results_between)
}
dev.off()

@@

# Passeriformes shapespace analyses

@@

family_shapespace <- MCMCglmm.subsets(n = 1000, # Just run it on 1k matrices for now
      data          = shapespace,
      posteriors    = shapespace_passeriformes,
      group         = MCMCglmm.levels(shapespace_passeriformes)[c(famil_lvl, phylo_lvl)],
      rename.groups = c(levels(shapespace$level2)[-1], "phylogeny")) #[-1] is because the first level is ""
## Clean garbage
rm(shapespace_passeriformes)

## Loading the results
load(file = "../Data/Processed/shapespace_passeriformes_suborder_family_results_list.rda")
results <- shapespace_passeriformes_suborder_family_results_list
rm(shapespace_passeriformes_suborder_family_results_list)
```

@@

```{r}
box_size <- 1.5
pdf(file = "../Manuscript/Figures/Figure1_passeriformes.pdf", height = box_size*7, width = box_size*10)

## Global window
lay_mat_family <- rbind(
        matrix(c(1,1, 2:9), ncol = 10, nrow = 1), # line 1
        matrix(c(1,1, 10:17), ncol = 10, nrow = 1), # line 2
        matrix(c(1,1, 18:25), ncol = 10, nrow = 1), # line 3
        matrix(c(26:35), ncol = 10, nrow = 1), # line 4
        matrix(c(36:45), ncol = 10, nrow = 1), # line 5
        matrix(c(46:55), ncol = 10, nrow = 1) # line 6
    )
lay_mat_sub <- rbind(
        matrix(c(1,1, 2:5, rep(0, 4)), ncol = 10, nrow = 1) # line 7
    )
heights_vals <- rep(1, 8)
widths_vals <- rep(c(1, 1/3), 5)
layout_matrix <- layout(rbind(lay_mat_sub, (lay_mat_family+5)), widths = widths_vals, heights = heights_vals, respect = TRUE)

## PLOTTING THE SUBORDERS

## Getting the tip order for the ladderised super order tree
tip_order <- c(
    "Tyranni" = 1,
    "Passeri" = 2)

## Plotting the tree and getting the tip orders
tip_colours_suborders <- plot.short.tree(tree, shapespace, "level1", tip.order = tip_order)
save(tip_colours_suborders, file = "../Data/Processed/tip_colours_suborders.rda")

## Plotting the ellipses
for(i in 1:length(tip_order)) {
    ## Plot each pannel
    wrap.plot.ellipses(i, tip_colours_suborders, suborder_shapespace, results$suborder_results_between)
}

## PLOTTING THE FAMILIES

@@

## Plotting the tree and getting the tip orders
tip_colours_families <- plot.short.tree(tree, shapespace, "level2", tip.order = tip_order)
save(tip_colours_families, file = "../Data/Processed/tip_colours_families.rda")

## Plotting the ellipses
for(i in 1:length(tip_order)) {
    ## Plot each pannel
    wrap.plot.ellipses(i, tip_colours_families, family_shapespace, results$family_results_between)
}
dev.off()
